Equalizer Programming Guide\\ {\footnotesize\mdseries \htmladdnormallink{http://www.equalizergraphics.com/documents/Developer/ProgrammingGuide.pdf} {http://www.equalizergraphics.com/documents/ProgrammingGuide.pdf}}

### 2007-09-21 Developer/ProgrammingGuide/paper.tex
Update Programming Guide and related illustrations.
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\title{Equalizer Programming Guide\\
  {\footnotesize\mdseries
    \htmladdnormallink{http://www.equalizergraphics.com/documents/Developer/ProgrammingGuide.pdf}
    {http://www.equalizergraphics.com/documents/ProgrammingGuide.pdf}}
}
\author{Stefan Eilemann\thanks{eile@eyescale.ch}\\[\medskipamount]
  % Eyescale Software Sarl
}
\date{
  \textbf{INCOMPLETE}\\[\medskipamount]
  Version 0.6, \today
}

@@

\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Version & \bf Date     & \bf Changes \\
    \hline
    0.6         & Sep 14, 2007 & added pipe, window and object manager\\
    0.5         & Sep 3,  2007 & added config and partly pipe\\
    0.4         & Aug 31, 2007 & added distributed objects\\
    0.3         & Aug 26, 2007 & added application and render client\\
    0.2         & Aug 20, 2007 & added main function\\
    0.1         & Aug 19, 2007 & outlined the basic concepts\\
    \hline
  \end{tabularx}}

\clearpage
\tableofcontents
\thispagestyle{empty}

\clearpage
\pagenumbering{arabic}

@@

The second step is to initialize the Equalizer library. The
initialization function of Equalizer also parses the command line, which
is used to set certain default values based on Equalizer-specific
options\footnote{Equalizer-specific options always start with --eq-},
e.g., the default server location. Furthermore, a node factory is
provided. The \textsf{EQERROR} macro, and its counterparts
\textsf{EQWARN}, \textsf{EQINFO} and \textsf{EQVERB} allow selective
debugging outputs with various logging levels:

@@

\begin{wrapfigure}{r}{.4\textwidth}
  \subfigure[]{\label{fSync}
    \includegraphics[width=.18\textwidth]{images/sync}}\hfil
  \subfigure[]{\label{fAsync}
    \includegraphics[width=.18\textwidth]{images/async}}%
  {\caption{\small\label{fSyncAsync}Synchronous and asynchronous execution}}
\end{wrapfigure}
\textsf{Config::finishFrame} synchronizes on the completion of the frame
\textsf{current - latency}. The latency is specified in the
configuration file, and allows several outstanding frames for which the
tasks are already queued in the node and pipe threads for
execution. This enables overlapped execution and minimizes idle
times. The first \textsf{latency Config::finishFrame} return
immediately, since they have no frame to synchronize
upon. Figure~\ref{fSyncAsync} shows the execution of (hypothetical)
rendering tasks without latency~(\ref{fSync}) and with a latency of
one~(\ref{fAsync}).

@@

{\footnotesize\begin{lstlisting}
const void* InitData::getInstanceData( uint64_t* size )
{
    *size = sizeof( uint32_t ) + sizeof( eq::WindowSystem ) +
            _filename.length() + 1;
    if( _instanceData )
        return _instanceData;

@@

    reinterpret_cast< uint32_t* >( _instanceData )[0] =  _frameDataID;
    reinterpret_cast< uint32_t* >( _instanceData )[1] =  _windowSystem;

    const char* string = _filename.c_str();
    memcpy( _instanceData + sizeof( uint64_t ), string, _filename.length()+1 );

@@

After the data is no longer needed, \textsf{releaseInstanceData} is
called by Equalizer to allow the object to free the data. Since the data
is cached for further usage, the release method is not overwritten.

@@

{\footnotesize\begin{lstlisting}
void InitData::applyInstanceData( const void* data, const uint64_t size )
{
    EQASSERT( size > ( sizeof( _frameDataID ) + sizeof( _windowSystem )));

    _frameDataID  = reinterpret_cast< const uint32_t* >( data )[0];
    _windowSystem = reinterpret_cast< const eq::WindowSystem* >( data )[1];
    _filename     = static_cast<const char*>( data ) + sizeof( uint64_t );

    EQASSERT( _frameDataID != EQ_ID_INVALID );

    EQINFO << "New InitData instance" << endl;
}
\end{lstlisting}}%>>

@@

Events are send by the render clients to the application using
{eq::Config::sendEvent}. At the end of the frame,
\textsf{Config::finishFrame} calls \textsf{Config::handleEvents} to do
the event handling. The default implementation processes all pending
events by calling \textsf{Config::handleEvent} for each of them.

@@

The \textsf{eqPly} example continuously renders new frames. It
implements \textsf{Config::hand\-le\-Event} to provide the various reactions
to user input:

@@

@@

    _model = PlyFileIO::read( filename.c_str( ));
    if( !_model)
        EQWARN << "Can't load model: " << filename << endl;

@@

\begin{wrapfigure}{r}{.4\textwidth}
  \includegraphics[width=.4\textwidth]{images/windowInit.pdf}
  {\caption{\small\label{fWindowInit}Window Initialization Flow Chart}}
\end{wrapfigure}

The initialization sequence uses multiple, overrideable task
methods. The main task method \textsf{configInit} executes a `child'
task method to create the drawable and context. The child task method
depends on the pipe's window system. The default implementations
\textsf{configInitGLX}, \textsf{configInitWGL} or \textsf{configInitAGL}
creates a visible window using OS-specific methods. If the OpenGL
drawable and context was created successfully, \textsf{configInit} calls
\textsf{configInitGL}, which performs the generic OpenGL state
initialization. The default implementation sets up some typical OpenGL
state, e.g., it enables the depth test.

@@

The \textsf{eqPly} window initialization function first calls
\textsf{eq::Window::configInit} to use the generic window setup. If that
was successful, it initializes an object manager and an overlay logo:

@@

    EQASSERT( !_objects );

    if( firstWindow == this )
    {
        _objects = new ObjectManager( getGLFunctions( ));
        _loadLogo();
    }
    else
    {
        _objects     = firstWindow->_objects;
        _logoTexture = firstWindow->_logoTexture;
        _logoSize    = firstWindow->_logoSize;
    }

    if( !_objects ) // happens if first window failed to initialize
        return false;

    return true;
}
\end{lstlisting}}

The object manager is used to create OpenGL objects in a multipipe,
multi-threaded execution environment. It is described in detail
below. It is used in conjunction with a reference pointer here, since it
is potentially `owned' by multiple windows at the same time.

@@

The window exit deallocates all OpenGL objects, if the object manager is
about to be disposed. The object manager does not delete the object in
its destructor, since it might now longer have a current OpenGL
context. In any case, \textsf{eq::Window::configExit} is called to
destroy the drawable and context:

{\footnotesize\begin{lstlisting}
bool Window::configExit()
{
    if( _objects.isValid() && _objects->getRefCount() == 1 )
        _objects->deleteAll();

    _objects = 0;
    return eq::Window::configExit();
}
\end{lstlisting}}

@@

@@

\if 0
{\footnotesize\begin{lstlisting}
\end{lstlisting}}
\fi

>>>>

\title{Equalizer Programming Guide\\
  {\footnotesize\mdseries
    \htmladdnormallink{http://www.equalizergraphics.com/documents/Developer/ProgrammingGuide.pdf}
    {http://www.equalizergraphics.com/documents/ProgrammingGuide.pdf}}
}
\author{Stefan Eilemann\thanks{eile@eyescale.ch}\\[\medskipamount]
  % Eyescale Software Sarl
}
\date{
  \textbf{INCOMPLETE}\\[\medskipamount]
  Version 0.7, \today
}

@@

\clearpage
\tableofcontents
\thispagestyle{empty}
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    0.7     & Sep 21, 2007 & started channel and image compositing,
                             added event handling, moved to new eqPly
                             code base\\
    0.6     & Sep 14, 2007 & added pipe, window and object manager\\
    0.5     & Sep 3,  2007 & added config and partly pipe\\
    0.4     & Aug 31, 2007 & added distributed objects\\
    0.3     & Aug 26, 2007 & added application and render client\\
    0.2     & Aug 20, 2007 & added main function\\
    0.1     & Aug 19, 2007 & outlined the basic concepts\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

\pagenumbering{arabic}

@@

The second step is to initialize the Equalizer library. The
initialization function of Equalizer also parses the command line, which
is used to set certain default values based on Equalizer-specific
options\footnote{Equalizer-specific options always start with -\,-eq-},
e.g., the default server location. Furthermore, a node factory is
provided. The \textsf{EQERROR} macro, and its counterparts
\textsf{EQWARN}, \textsf{EQINFO} and \textsf{EQVERB} allow selective
debugging outputs with various logging levels:

@@

\begin{wrapfigure}{r}{.6\textwidth}
  \subfigure[]{\label{fSync}
    \includegraphics[width=.25\textwidth]{images/sync}}\hfil
  \subfigure[]{\label{fAsync}
    \includegraphics[width=.25\textwidth]{images/async}}%
  {\caption{\small\label{fSyncAsync}Synchronous and asynchronous execution}}
\end{wrapfigure}
\textsf{Config::finishFrame} synchronizes on the completion of the frame
\textsf{current - latency}. The latency is specified in the
configuration file, and allows several outstanding frames for which the
tasks are already queued in the node and pipe threads for
execution. This enables overlapped execution and minimizes idle
times. The first \textsf{latency Config::finishFrame} return
immediately, since they have no frame to synchronize
upon. \fig{fSyncAsync} shows the execution of (hypothetical)
rendering tasks without latency~(\fig{fSync}) and with a latency of
one~(\fig{fAsync}).

@@

{\footnotesize\begin{lstlisting}
const void* InitData::getInstanceData( uint64_t* size )
{
    *size = sizeof( StaticData ) + _filename.length() + 1;
    if( _instanceData )
        return _instanceData;

@@

    reinterpret_cast< StaticData* >( _instanceData )[0] = _data;

    const char* string = _filename.c_str();
    memcpy( _instanceData + sizeof( StaticData ), string, _filename.length()+1);

@@

After the data is no longer needed, \textsf{releaseInstanceData} is
called by Equalizer to allow the object to free the data. Since the data
is cached for further usage, the release method is not overwritten in
\textsf{InitData}.

@@

{\footnotesize\begin{lstlisting}
void InitData::applyInstanceData( const void* data, const uint64_t size )
{
    _data     = reinterpret_cast< const StaticData* >( data )[0];
    _filename = static_cast<const char*>( data ) + sizeof( StaticData );

    EQASSERT( _data.frameDataID != EQ_ID_INVALID );
    EQINFO << "New InitData instance" << endl;
}
\end{lstlisting}}%>>

@@

Events are send by the render clients to the application using
\textsf{eq::Config::sendEvent}. At the end of the frame,
\textsf{Config::finishFrame} calls \textsf{Config::handleEvents} to do
the event handling. The default implementation processes all pending
events by calling \textsf{Config::handleEvent} for each of them.

@@

The \textsf{eqPly} example continuously renders new frames. It
implements \textsf{Config::hand\-le\-Event} to provide the various reactions
to user input, most importantly camera updates based on mouse
events. The camera position has to be handled correctly with respect to
latency, and is therefore saved in the frame data:

@@

        case eq::ConfigEvent::POINTER_MOTION:
            if( event->pointerMotion.buttons == eq::PTR_BUTTON_NONE )
                return true;

            if( event->pointerMotion.buttons == eq::PTR_BUTTON1 )
            {
                _spinX = 0;
                _spinY = 0;

                _frameData.data.rotation.preRotateX( 
                    -0.005f * event->pointerMotion.dx );
                _frameData.data.rotation.preRotateY(
                    -0.005f * event->pointerMotion.dy );
            }
            else if( event->pointerMotion.buttons == eq::PTR_BUTTON2 ||
                     event->pointerMotion.buttons == ( eq::PTR_BUTTON1 |
                                                       eq::PTR_BUTTON3 ))
            {
                _frameData.data.translation.z +=
                    .005f * event->pointerMotion.dy;
            }
            else if( event->pointerMotion.buttons == eq::PTR_BUTTON3 )
            {
                _frameData.data.translation.x += 
                    .0005f * event->pointerMotion.dx;
                _frameData.data.translation.y -= 
                    .0005f * event->pointerMotion.dy;
            }
            return true;

@@

    _model = new Model();
    if ( !_model->readFromFile( filename.c_str() ) )
    {
        EQWARN << "Can't load model: " << filename << endl;
        delete _model;
        _model = 0;
    }

@@

\begin{wrapfigure}{r}{.4\textwidth}
  \includegraphics[width=.4\textwidth]{images/windowInit.pdf}
  {\caption{\small\label{fWindowInit}Window Initialization}}
\end{wrapfigure}

The initialization sequence uses multiple, overrideable task
methods. The main task method \textsf{configInit} executes a `child'
task method to create the drawable and context. The child task method
depends on the pipe's window system. The default implementations
\textsf{configInitGLX}, \textsf{configInitWGL} or \textsf{configInitAGL}
create an on-screen window using OS-specific methods. If the OpenGL
drawable and context was created successfully, \textsf{configInit} calls
\textsf{configInitGL}, which performs the generic OpenGL state
initialization. The default implementation sets up some typical OpenGL
state, e.g., it enables the depth test.

@@

The \textsf{eqPly} window initialization function first calls
\textsf{eq::Window::configInit} to use the generic window setup. If that
was successful, it initializes a state object and an overlay logo:

@@

    EQASSERT( !_state );

    if( firstWindow == this )
    {
        _state = new VertexBufferState( getGLFunctions( ));
        _loadLogo();

        const Node*     node     = static_cast< const Node* >( getNode( ));
        const InitData& initData = node->getInitData();

        if( initData.useVBOs() )
        {
            const eq::GLFunctions* glFunctions = getGLFunctions();
            // Check if all VBO funcs available, else leave DISPLAY_LIST_MODE on
            if( glFunctions->hasGenBuffers() && glFunctions->hasBindBuffer() &&
                glFunctions->hasBufferData() && glFunctions->hasDeleteBuffers())
            {
                _state->setRenderMode( mesh::BUFFER_OBJECT_MODE );
                EQINFO << "VBO rendering enabled" << endl;
            }
            else
                EQWARN << "VBO function pointers missing, using display lists" 
                       << endl;
        }
    }
    else
    {
        _state       = firstWindow->_state;
        _logoTexture = firstWindow->_logoTexture;
        _logoSize    = firstWindow->_logoSize;
    }

    if( !_state ) // happens if first window failed to initialize
        return false;

The state object is used to handle the creation of OpenGL objects in a
multipipe, multi-threaded execution environment. It uses an object
manager, which is described in detail below. It is used in conjunction
with a reference pointer here, since it is potentially `owned' by
multiple windows at the same time.

@@

The window exit deallocates all OpenGL objects if the state object is
about to be disposed. The object manager does not delete the object in
its destructor, since it does not know if an OpenGL context is still
current. In any case, \textsf{eq::Window::configExit} is called to
destroy the drawable and context:

{\footnotesize\begin{lstlisting}
bool Window::configExit()
{
    if( _state.isValid() && _state->getRefCount() == 1 )
        _state->deleteAll();

    _state = 0;
    return eq::Window::configExit();
}
\end{lstlisting}}

@@

The state object in \textsf{eqPly} gathers all rendering state, which
includes an object manager for object state handling. This state object
can easily be replaced to use the rendering code in a stand-alone
application, e.g., for an GLUT-based renderer.

@@

During channel initialization, the near and far planes are set to
reasonable values to contain the whole model. During rendering, the near
and far planes are adjusted to the current model position:

{\footnotesize\begin{lstlisting}
bool Channel::configInit( const uint32_t initID )
{
    setNearFar( 0.1f, 10.0f );
    return true;
}
\end{lstlisting}}

\subsubsection{Rendering}

The central rendering routine is \textsf{Channel::frameDraw}. This
routine contains the application's OpenGL rendering code, which is being
rendered using the rendering context provided by Equalizer.

In \textsf{eqPly}, the OpenGL context is first set up using various
\textsf{apply} convenience methods from the base Equalizer channel
class. Each of the \textsf{apply} methods uses the corresponding
\textsf{get} method(s) and then calls the approriate OpenGL
function(s). It is also possible to just query the values from Equalizer
using the \textsf{get} methods, and use them to set up the OpenGL state
appropriatly, for example by passing the parameters to the renderer used
by the application.

For example, the implementation for \textsf{eq::Channel::applyBuffer}
does set up the correct rendering buffer and color mask, which depends
on the current eye pass and possible anaglyphic stereo parameters:

{\footnotesize\begin{lstlisting}
void eq::Channel::applyBuffer()
{
    glReadBuffer( getReadBuffer( ));
    glDrawBuffer( getDrawBuffer( ));

    const ColorMask& colorMask = getDrawBufferMask();
    glColorMask( colorMask.red, colorMask.green, colorMask.blue, true );
}
\end{lstlisting}}

The following contextual information has to be used in order to render
the view expected by Equalizer:
\begin{description}
\item[Buffer]: The OpenGL read and draw buffer as well as color mask.
  These parameters are influenced by the current eye pass, eye
  seperation and anaglyphic stereo settings.
\item[Viewport]: The two-dimensional pixel viewport restricting the
  rendering area within the channel. For correct operations, both
  \textsf{glViewport} and \textsf{glScissor} have to be used. The pixel
  viewport is influenced by the destination channel's viewport
  definition and compound viewports set for sort-first/2D decompositions.
\item[Frustum]: Frustum parameters as defined by
  \textsf{glFrustum}. The frustum is influenced by the destination
  channel's view definition, compound viewports, head matrix and the
  current eye pass.
\item[Head Transformation:] A transformation matrix positioning the
  frustum. This is typically an identity matrix and is used for off-axis
  frustra in immersive rendering.
\item[Range]: A one-dimensional range with the interval [0..1]. This
  parameter is optional and should be used by the application to render
  only the appropriate subset of its data.
\end{description}

Consequently, the first part of the \textsf{frameDraw} method in
\textsf{eqPly} look as follows:

{\footnotesize\begin{lstlisting}
void Channel::frameDraw( const uint32_t frameID )
{
    applyBuffer();
    applyViewport();

    glMatrixMode( GL_PROJECTION );
    glLoadIdentity();

    applyFrustum();

    glMatrixMode( GL_MODELVIEW );
    glLoadIdentity();
    applyHeadTransform();
\end{lstlisting}}

\begin{wrapfigure}{r}{.4\textwidth}
  \includegraphics[width=.4\textwidth]{images/DB.pdf}
  {\caption{\small\label{fDB}Final view of an DB compound}}
\end{wrapfigure}

After the basic view setup, a directional light is set up, and the model
is positioned using the camera parameters from the frame data. The
camera parameters are transported using the the frame data to ensure
that all channels render the frame from the same position.

Furthermore, a color is set in case the model does not contain color
information, or the color information is not to be used. In sort-last
rendering, \textsf{eqPly} renders the individual parts with different
colors to illustrate the algorithm, as shown in \fig{fDB}. The Equalizer
channel provides a method to obtain a random, but unique color for all
channels in the configuration:

{\footnotesize\begin{lstlisting}
    glLightfv( GL_LIGHT0, GL_POSITION, lightpos );

    const Pipe*      pipe      = static_cast<Pipe*>( getPipe( ));
    const FrameData& frameData = pipe->getFrameData();

    glTranslatef( frameData.data.translation.x,
                  frameData.data.translation.y,
                  frameData.data.translation.z );
    glMultMatrixf( frameData.data.rotation.ml );

    Node*            node  = (Node*)getNode();
    const Model*     model = node->getModel();
    const eq::Range& range = getRange();

    if( !range.isFull( )) // Color DB-patches
    {
        const vmml::Vector3ub color = getUniqueColor();
        glColor3ub( color.r, color.g, color.b );
    }
    else if( !frameData.data.color || (model && !model->hasColors( )) )
    {
        glColor3f( 1.0f, 1.0f, 1.0f );
    }

\end{lstlisting}}

Finally the model is rendered, if it could be loaded during
initialization. Otherwise a quad is drawn in its place:

{\footnotesize\begin{lstlisting}
    if( model )
    {
        _drawModel( model );
    }
    else
    {
        glColor3f( 1.f, 1.f, 0.f );
        glNormal3f( 0.f, -1.f, 0.f );
        glBegin( GL_TRIANGLE_STRIP );
        glVertex3f(  .25f, 0.f,  .25f );
        glVertex3f(  .25f, 0.f, -.25f );
        glVertex3f( -.25f, 0.f,  .25f );
        glVertex3f( -.25f, 0.f, -.25f );
        glEnd();
    }
}
\end{lstlisting}}

\if 0
{\footnotesize\begin{lstlisting}
void Channel::_drawModel( const Model* model )
{
    Window*                  window    = static_cast<Window*>( getWindow() );
    mesh::VertexBufferState& state     = window->getState();

    const Pipe*              pipe      = static_cast<Pipe*>( getPipe( ));
    const FrameData&         frameData = pipe->getFrameData();

    const eq::Range&         range     = getRange();
    vmml::FrustumCullerf     culler;

    state.setColors( frameData.data.color && 
                     range.isFull() && 
                     model->hasColors() );
    _initFrustum( culler, model->getBoundingSphere( ));

    model->beginRendering( state );

\end{lstlisting}}

{\footnotesize\begin{lstlisting}
\end{lstlisting}}

The model data is organized a in 3-dimensional kD-tree\footnote{See also
  \link{http://en.wikipedia.org/wiki/Kd-tree}} for efficient view
frustum culling. During preprocessing, each node of the tree gets a
database range assigned.

{\footnotesize\begin{lstlisting}
\end{lstlisting}}
\fi

\section{Advanced Features}

This section discusses some features of interest which are not covered
by the \textsf{eqPly} section. Where possible, code examples from the
Equalizer distribution are used.

\subsection{Event Handling}

Event handling is one of the areas which requires a lot of
flexibility. On one hand, the implementation differs slightly for each
window system due to conceptual differences in the OS-specific
implementation. On the other hand, each application and widget set has
its own model on how events are to be handled. Therefore, event handling
in Equalizer is customizable in any stage of the processing\footnote{\link{http://www.equalizergraphics.com/documents/design/eventHandling.html}}.

The default implementation aims to provide a quick and easy
implementation of event handling, while allowing all necessary
customizations. It gathers all events in the application main thread, so
that the developer only has to implement \textsf{Config::processEvent}
to update its data based on the pre-processed, generic events.

\subsubsection{Threading}

Where possible, events are received and processed by a separate per-node
event thread to allow asynchronous event handling. Currently an event
thread is only used by the X11/glX window system. WGL receives and
processes the events from the pipe threads which created the
windows. AGL receives the events from application or node main
thread. Whenever the term \textbf{event thread} is used, it refers to
the thread receiving the event, i.e., a per-node thread for glX, the pipe
thread for WGL and the main thread for AGL.

\begin{wrapfigure}{r}{.4\textwidth}
  \includegraphics[width=.4\textwidth]{images/eventInit.pdf}
  {\caption{\small\label{fEventInit}Event Handling Initialization}}
\end{wrapfigure}

During window and pipe initialization the event handling is set up. For
both entities, \textsf{initEvent\-Handler} is called to register the pipe
or window with an event handler. This method may be overwritten to use a
custom event handler, or to not install an event handler to disable
event handling in Equalizer. Likewise, \textsf{exitEventHandler} is
called to de-initialize event handling.

An event handler consists of two parts: the generic base class providing
the interface and some generic functions, and the window-system-specific
part providiing the actual implementation.

Pipe event handling is only used for glX, where one \textsf{Display}
connection is created to subscribe to window events. Event handling is
initialized whenever a new, window-system-specific pipe or window handle
is set. First, \textsf{exitEventHandler} is called to de-initialize
event handling for the old handle (if set), and then
\textsf{initEvent\-Handler} is called for the new handle. AGL and glX use
an event handler singleton, whereas WGL uses one event handler per
window.

\subsubsection{Message Pump}

WGL and AGL require the application to manually receive and dispatch
(`pump') events. On WGL, this has to happen on each thread with windows,
whereas on AGL it has to happen only on the main thread. By default,
Equalizer pumps these events automatically for the application.

The methods \textsf{Client::useMessagePump} and
\textsf{Pipe::useMessagePump} can be overriden to return \textsf{false}
to disable this behaviour for their respective threads. On non-threaded
pipes, \textsf{Pipe::useMessagePump} is not called.

\subsubsection{Event Data Flow}

\begin{wrapfigure}{r}{.4\textwidth}
  \includegraphics[width=.4\textwidth]{images/eventHandling.pdf}
  {\caption{\small\label{fEventProcessing}Event Processing}}
\end{wrapfigure}

Events are received by an event handler. The event handler finds the
\textsf{eq::Window} associated to the event. It then creates a generic
\textsf{WindowEvent}, which holds important event data in a format
independent of the window system. The original event is attached to the
generic window event.

The event handler then passes the window event to
\textsf{Window::processEvent}, which is responsible for either handling
the event locally, or for translating it into a generic
\textsf{ConfigEvent}. The config events are send to the application
thread using \textsf{Config::sendEvent}. If the event was processed, the
function has to return \textsf{true}. If \textsf{false} is returned, the
event will be passed to a previously installed, window-system-specific
event handling function. The default implementation of
\textsf{Window::processEvent} passes most events on to the application.

Events sent using \textsf{Config::sendEvent} are queued in the
application thread. After a frame has been finished,
\textsf{Config::finishFrame} calls \textsf{Config::processEvents}. The
default implementation of this method provides non-blocking event
processing, that is, it calls \textsf{Config::processEvent} for each
queued event. By overriding this function, event-driven execution can be
implemented.

\subsubsection{Custom Events in eqPixelBench}

The \textsf{eqPixelBench} example is a benchmark program to measure the
pixel transfer rates from and to the framebuffer of all channels within
a configuration. It uses custom config events to send the gathered data
to the application. It is much simpler than the \textsf{eqPly} example
since it does not provide any useful rendering or user interaction.

The rendering routine of \textsf{eqPixelBench} in
\textsf{Channel::frameDraw} loops through a number of pixel formats and
types. For each of them, it measures the time to readback and assemble a
full-channel image. The format, type, size and time is recorded in a
config event, which is sent to the application.

The \textsf{ConfigEvent} derives from the \textsf{eq::ConfigEvent}
structure and has the following definition:

{\footnotesize\begin{lstlisting}
struct ConfigEvent : public eq::ConfigEvent
{
public:
    enum Type
    {
        READBACK = eq::ConfigEvent::USER,
        ASSEMBLE
    };

    ConfigEvent()
        {
            size = sizeof( ConfigEvent );
        }

    // channel name is in user event data
    char           formatType[64];
    vmml::Vector2i area;
    float          msec;
};
\end{lstlisting}}

The \textsf{Config::sendEvent} method transmits a
\textsf{eq::ConfigEvent} or derived class to the application. The
ConfigEvent has to be a C-type structure, and its \textsf{size}
member has to be set to the full size of the event to be transmitted.
Each event has a type, by which the config processing function can
identify it.

User-defined types start at \textsf{eq::ConfigEvent::USER}, and the
member variable \textsf{ConfigEvent::user} can be used to store
\textsf{EQ\_USER\_EVENT\_SIZE}\footnote{currently 32 bytes} bytes. In
this space, the channel's name is stored. Additional variables are used
to transport the pixel format and type, the size and the time it took
for rendering.

On the application end, \textsf{Config::handleEvent} uses the ostream
operator for the derived config event to output these events in a nicely
formatted way:

{\footnotesize\begin{lstlisting}
std::ostream& operator << ( std::ostream& os, const ConfigEvent* event );
...
bool Config::handleEvent( const eq::ConfigEvent* event )
{
    switch( event->type )
    {
        case ConfigEvent::READBACK:
        case ConfigEvent::ASSEMBLE:
            cout << static_cast< const ConfigEvent* >( event ) << endl;
            return true;

        default:
            return eq::Config::handleEvent( event );
    }
}
\end{lstlisting}}%>>

\subsection{Image Compositing for Scalable Rendering}

Two task methods are responsible for collecting and compositing the
result image during scalable rendering, when multiple channels are
rendering for a single view. The source channels producing an
\textsf{outputFrame} use \textsf{Channel::frameReadback} to read the
pixel data from the frame buffer. The channels receiving one or multiple
\textsf{inputFrame} use \textsf{Channel::frameAssemb\-le} to assemble
the frames into the framebuffer. Equalizer takes care of the network
transport of frame buffer data between nodes, if needed.

\subsubsection{Frame, FrameData and Images}

A \textsf{eq::Frame} references a \textsf{eq::Fra\-me\-Data}. The frame data
is the object connecting output with input frames. It is a holder for
images and is used to signal the availability of the pixel data. An
\textsf{eq::Image} holds a two-dimensional snapshot of the framebuffer
and can contain color and/or depth information.

Readback and assemble operations on frames and images are designed to be
asynchronous. They have a start and finish method for readback and
assemble to allow the initiation and synchronization of the operation.
Currently, only synchronous readback and assembly using
\textsf{glReadPixels} and \textsf{glDrawPixels} is implemented in the
respective start method of the image.

\if 0

\begin{wrapfigure}{r}{.6\textwidth}
  \includegraphics[width=.6\textwidth]{images/assembly.pdf}
  {\caption{\small\label{fAssembly}Hierarchy of assembly classes}}
\end{wrapfigure}

The offset of input and output frames characterizes the position of the
frame data with respect to the framebuffer, that is, the
\textbf{window's} lower-left corner. For output frames this is the
position of the channel in 
\fi

\subsubsection{Custom Assembly in eVolve}

\subsection{Head Tracking}
\if 0
{\footnotesize\begin{lstlisting}
\end{lstlisting}}
\fi
